---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading required R packages to peform the analysis

```{r}
library(timeDate)
library(plyr)
library(sqldf)
library(lattice)
```

## Loading and preprocessing the data

```{r}
# downloading and unzipping the file in the working directory
dataUrl <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(dataUrl, destfile = "data.zip", mode = "wb")
unzip("data.zip")
unlink("data.zip")
dateDownloaded <- date()
originalData <- read.csv("activity.csv", header=TRUE)


activity <- originalData
activity$date <- as.Date(activity$date)
activity$day <- as.factor(weekdays(activity$date))
activity$interval <- as.factor(activity$interval)
activity$week <- as.factor(ifelse(weekdays(activity$date) %in% c("Saturday","Sunday"), 'Weekend', 'Weekday'))

activity_NAomit <- na.omit(activity)
```


## What is mean total number of steps taken per day?

#### 1) The total number of steps per day (ignoring NAs) is:
```{r}
activity_NAomit_perDay <- ddply(activity_NAomit,.(date),summarise,steps=round(sum(steps),0))
activity_NAomit_perDay
```


#### 2) Histogram of the total number of steps taken each day (ignoring NAs):
```{r}
hist(activity_NAomit_perDay$steps, breaks=15, xlab="Number of Steps Per Day",ylab="Frequency", main="Histogram")
```


#### 3) The average mean and median of steps per day (ignoring NAs) is, respectively:
```{r}
mean(activity_NAomit_perDay$steps)
```

```{r}
median(activity_NAomit_perDay$steps)
```

## What is the average daily activity pattern?

#### 1) Time series plot of the average of steps per interval across all days.

```{r}
activity_NAomit_perInterval <- ddply(activity_NAomit,.(interval),summarise,steps=round(mean(steps),2))
plot(as.numeric(activity_NAomit_perInterval$interval),activity_NAomit_perInterval$steps, type="l", xaxt="n", xlab="Interval",ylab="Average Steps Per Interval", main="Time Series Plot Of The Average Steps Per Interval")
```

#### 2) The interval 835 scored the highest average of number of steps: 206.17.
```{r}
head(activity_NAomit_perInterval[order(-activity_NAomit_perInterval$steps),])
```


## Imputing missing values

#### 1) The total number of rows with missing values (NA) is 2304. The only variable with missing values is: steps.

```{r}
sum(is.na(activity))
summary(activity)
```

#### 2) In order to adress the missing values issue in the steps variable I will use the average of steps of the interval given the week variable (considering it might be a difference in the pattern when is a weekday or weekend).

 * Estimate the average of steps by interval and weekend/weekday.
 * Create a variable "key" in the activity data in order to join datasets.
 * Run a for loop to fill in NA values. 


#### 3) Creating a new dataset activity_new

```{r}
activity_new <- activity

stepsFix <- ddply(activity_NAomit,.(interval,week),summarise,stepsMean=round(mean(steps),2))
stepsFix$key <- paste(stepsFix$interval,stepsFix$week,sep="")
stepsFix <- stepsFix[c(1,2,4,3)]

activity_new$key <- paste(activity_new$interval,activity_new$week,sep="")
activity_new$isNA <- is.na(activity_new$steps)
activity_new <- sqldf("SELECT activity_new.*, stepsFix.stepsMean FROM activity_new INNER JOIN stepsFix ON activity_new.key = stepsFix.key")
for(i in 1:length(activity_new$steps)){ if(is.na(activity_new$steps[i])){   activity_new$steps[i] <- activity_new$stepsMean[i]}}

activity_new_perDay <- ddply(activity_new,.(date),summarise,steps=round(sum(steps),0))
activity_new_perInterval <- ddply(activity_new,.(interval),summarise,steps=round(mean(steps),2))

```

#### 4) After considering average of steps for the missing values the mean and median is sliglty different. In the first analyis the mean and median was nearly equal, however in this part they are more distant which also indicate a right skewed pattern (given the mean is higher than the median).

```{r}
hist(activity_new_perDay$steps, breaks=15, , xlab="Number of Steps Per Day (Average)",ylab="Frequency", main="Histogram")

mean(activity_new_perDay$steps)

median(activity_new_perDay$steps)
```


## Are there differences in activity patterns between weekdays and weekends?

#### 1) The factor Weekend/Weekday was created in the first task of the assigment.

#### 2) Panel comparing the average of steps pattern in Weekend and Weekday. 


```{r}
activity_perWeek <- ddply(activity_new,.(interval,week),summarise,stepsMean=round(mean(steps),2))
xyplot(stepsMean~interval|week, type="l", data=activity_perWeek, as.table=FALSE, layout = c(1, 2))

#Weekend
sum(subset(activity_perWeek, activity_perWeek$week == "Weekend")$steps)

#Weekday
sum(subset(activity_perWeek, activity_perWeek$week == "Weekday")$steps)
```


From the plot and the total sum of steps it's possible to deduct that:

 * The subject starts walking earlier on Weekdays.
 * The subject walks more on Weekends.


